Human capital mediates natural selection in contemporary humans

### 2022-01-03 why-natural-selection.Rmd
Mention weights in tables and figures.

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more
among low-income individuals. However, our effect sizes appear unchanged
over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
\@ref(tab:tbl-children-by-period)), during which
government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter $a$
in the model [cf. @jones2008fertility]. This theory would not explain why our
results are weaker at higher incomes and education levels.[^note] A third
alternative is that traits under selection are linked to externalizing behaviour
and risk-seeking. This might be partially captured by our risk-aversion
parameter $\sigma$, which affects fertility by changing the marginal utility of
income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of
children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
33 regressions). Given that, and the statistical power we get from our sample
size, we believe that the non-result is real: while risk attitude does predict
fertility in the sample, it is not an important channel for natural selection.

[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$,
this effect becomes stronger, not weaker, at higher wages.

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents."}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings).")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children).")
```

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.", fig.align = "center"}

@@

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on *number of siblings* on these
subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
by terciles of age at first live birth, for mothers and fathers. As in the
respondents' generation, effect sizes are larger for younger parents. This
holds for both sexes.

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth."}

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings."}

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}

@@

```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data. Each dot represents one bivariate regression. Absolute effect sizes are plotted. Points are jittered on the Y axis. Top principal components are labelled."}

>>>>

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, our effect sizes appear unchanged over a large
span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period)), during which government spending on
child-related benefits varied considerably [@hoc1999socsec]. In general, there
is only weak evidence that welfare benefits affect fertility
[@gauthier2007impact]. A second alternative theory is that polygenic scores
correlate with the motivation to have children, i.e. parameter $a$ in the model
[cf. @jones2008fertility]. This theory would not explain why selection effects
are smaller at higher incomes and education levels.[^note] A third alternative
is that traits under selection are linked to externalizing behaviour and
risk-seeking. This might be partially captured by our risk-aversion parameter
$\sigma$, which affects fertility by changing the marginal utility of income,
but a more direct channel is risky sexual behaviour [@mills2021identification].
The data here provide some support for this: scores which might plausibly be
linked to externalizing behaviour, like ADHD and younger age at smoking
initiation, are selected for. However, this theory is less good at explaining
variation in selection across the full range of scores, including physical
measures such as waist-hip ratio and BMI. We test this theory directly by
re-estimating equation \@ref(eq:regression) controlling for a measure of risk
attitude (UK Biobank field 2040).  The median ratio of effect sizes between
regressions with and without controls is `r median(ratio)`; all scores which are
significant at $p < 0.05/33$ in uncontrolled regressions remain so when
controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.
[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$, this
effect becomes stronger, not weaker, at higher wages.

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents. Age/qualification weights. Parental generation weights multiplied by 1/number of siblings. Respondents' regression excludes childless respondents."}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings). Weights: Age/Qualification times 1/number of siblings.")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children). Weights: Age/Qualification.")
```

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area. Higher = more deprived. Estimates weighted by 1/number of siblings.", fig.align = "center"}

@@

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on *number of siblings* on these
subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
by terciles of age at first live birth, for mothers and fathers. As in the
respondents' generation, effect sizes are smaller, or even oppositely signed, for older parents. Importantly, this holds for both sexes.

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.  Estimates weighted by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Estimates weighted by 1/number of siblings."}

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Estimates weighted by 1/number of siblings."}

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings."}

@@

```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data. Each dot represents one bivariate regression. Sibling estimates weighted by 1/number of siblings. Absolute effect sizes are plotted. Points are jittered on the Y axis. Top principal components are labelled."}
